Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-03-25 manuscript.tex
Discussion finished.

The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
within one nucleosome depleted regions (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
TWIST1 and ALX4 heterodimer in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest. For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances. Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (SUPPLEMENT FIG). Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. Similarly, we observed that FOXA2 binding sites that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity. The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.

Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancer
consist out of clusters of 4-5 different TFBS. Homeotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain. Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}. It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}. It has been hypothesized that specificity is mainly achieved through cooperative binding with co-factors\cite{mannChapter3Hox2009}. Indeed, we observe this for EN1 and HOXB3 that are predicted to co-bind with PAX8 and X respectively forming specific dimer motifs. In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for TFX and TFY that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites. Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity. In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that can be important for cell type specific binding.

@@

%% clusters of low affinity HOX binding sites
% https://www.cell.com/cell/fulltext/S0092-8674(14)01518-9
% https://elifesciences.org/articles/28975

@@

% might be interesting
% https://www.nature.com/articles/s41594-024-01449-6
% https://www.nature.com/articles/s41586-018-0549-5

% co-factors (limitation)
% "new" studies
% https://www.biorxiv.org/content/10.1101/2022.08.26.505496v1.abstract
% https://www.cell.com/cell/fulltext/S0092-8674(20)31541-5?uuid=uuid%3A51e14242-dce3-432c-8521-14d9a94b0381
% https://www.cell.com/molecular-cell/fulltext/S1097-2765(21)01070-4?uuid=uuid%3Afbec3fc1-2707-4902-bda8-f196da451272
% https://www.science.org/doi/full/10.1126/science.adf6149
% focus on single enhancers and not entire locus
%
% from lambert et al
%%TFs have traditionally been classified as either ‘‘activators’’ or ‘‘repressors’’; however, this notion has been repeatedly questioned. Many TFs can recruit multiple cofactors that have opposite effects (Frietze and Farnham, 2011; Rosenfeld et al., 2006;
% Schmitges et al., 2016), dependent on the local sequence
%context and availability of cofactors (Meijsing et al., 2009;
%Wong and Struhl, 2011).

@@

>>>>

The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
within one nucleosome depleted regions (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
TWIST1 and ALX4 heterodimer in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest. For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances. Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (SUPPLEMENT FIG). Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity. The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par

Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancer
consist out of clusters of 4-5 different TFBS. Homeotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain. Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}. It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}. It has been hypothesized that specificity is mainly achieved through cooperative binding with other TFs\cite{mannChapter3Hox2009}. Indeed, we observe this for EN1 and HOXB3 that are predicted to co-bind with PAX8 and X respectively forming specific dimer motifs. In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for TFX and TFY that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites. Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity. In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that could be important for cell type specific binding.\par

In conclusion, we provide an enhancer-code atlas of early human neuronal and neural crest development. We confirm the
paradigm that enhancers consist out of clusters of TFBS with soft syntactic rules but also provide a counter example to
this paradigm. Furthermore, we validate the biological relevance of neural tube organoids. Finally, we provide $DeepNeuralTube$, two sequence-to-chromatin accessibility models, freely to the research community, which can be used to predict, interpret and design enhancers underlying early human neuronal and neural crest development.

@@

@@

@@

In this study we focused on predicting TF binding to enhancers, but not so much on their effector function. The effect a
TF has on target gene expression depends on its effector domain and interactions with
co-factors\cite{lambertHumanTranscriptionFactors2018} and this may depend on the sequence context of
the enhancer itself\cite{jacobsWidespreadRegulatorySpecificities2023}. Large scales studies, in a variety of cell
states, are now being performed to map TFs to co-factors and downstream effects on gene
expression\cite{tyckoHighThroughputDiscoveryCharacterization2020, alerasoolIdentificationFunctionalCharacterization2022,
sotoCompendiumHumanTranscription2022, delrossoLargescaleMappingMutagenesis2023,
mukundHighthroughputFunctionalCharacterization2023, bellComparativeCofactorScreens2024,
tyckoDevelopmentCompactTranscriptional2024, doughtySinglemoleculeStatesLink2024}. These studies may especially be
important to infer repressive roles of TFs on gene expression, an effect that is currently difficult to study with state
of the art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
dewinterModellingDesignTranscriptional2025}. In particular for the neural tube, it is known that repressive interactions
in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. In this study we could only predict a strong
repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024} (SUPPLEMENT FIG).\par

Furthermore, the expression of a gene is in general regulated by multiple enhancers in the gene's locus\cite{perryShadowEnhancersFoster2010,
martinez-araSystematicAnalysisIntrinsic2022, bergmanCompatibilityRulesHuman2022, zuinNonlinearControlTranscription2022,
gschwindEncyclopediaEnhancergeneRegulatory2023, martinez-araLargescaleAnalysisIntegration2024,
loubiereDevelopmentalHousekeepingTranscriptional2024, thomasEnhancerCooperativityCan2025}. In this study we focused on
the enhancer code of individual enhancers but not how the signal of multiple enhancers is integrated to control target
gene expression. Currently efforts are ongoing to build large scale models for modeling gene expression making use of the
code of all enhancers in the gene's locus\cite{avsecEffectiveGeneExpression2021,linderPredictingRNAseqCoverage2025a,
hingerlScoobyModelingMultimodal2024, hingerlFlashzoiEnhancedBorzoi2024, lalDecodingSequenceDeterminants2025}.\par

To conclude, we suspect that future models will need to take into account both the effector function of TFs through
modeling co-factor interactions and the entire gene's locus (including all enhancers and promoter(s)) to effectively
model GRNs underlying the cell's transcriptome.
